Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-10-11 manuscript/main.tex
incorporated Table 5 for subscales

\begin{abstract}
The gastrointestinal tract, its resident microorganisms, and the central
nervous system are connected by biochemical signaling, also known as the
"microbiome-gut-brain-axis." Both the human brain and the gut microbiome
have critical developmental windows in the first years of life,
raising the possibility that their development is co-occurring and
likely co-dependent. Emerging evidence implicates gut microorganisms and
microbiota composition in cognitive outcomes and neurodevelopmental
disorders (e.g., autism and anxiety), but the influence of gut microbial
metabolism on typical neurodevelopment has not been explored in detail.
We investigated the relationship of the microbiome with the neuroanatomy
and cognitive function of 381 healthy children, demonstrating that
differences in gut microbial taxa and gene functions are associated with
overall cognitive function and with differences in the size of
multiple brain regions.
Using a combination of multivariate linear and machine learning (ML) models,
we showed that many species,
including \emph{Alistipes obesi} and \emph{Blautia wexlerae}, were
associated with higher cognitive function, while some
species such as \emph{Ruminococcus gnavus} were more commonly found in
children with low cognitive scores after controlling for sociodemographic factors.
Microbial genes for enzymes involved in the metabolism of
neuroactive compounds, particularly short-chain fatty acids such as
acetate and propionate, were also associated with cognitive function.
In addition, ML models were able to use microbial taxa to predict the
volume of brain regions, and many taxa that were identified
as important in predicting cognitive function also dominated the
feature importance metric for individual brain regions,
and for specific subscales of cognitive function.
For example, \emph{B. wexlerae}
was the most important species in models predicting the size of the parahippocampal region
in both the left and right hemispheres
and was among the top predictors of gross motor and expressive language performance.
Several species from
the phylum \textit{Bacteroidetes}, including GABA-producing
\emph{B. ovatus}, were important for predicting
the size of the left accumbens area, but not the right. These
findings provide potential biomarkers of neurocognition and brain development
and may lead to the future development of targets for early detection and early
intervention.
\end{abstract}

@@

The first years of life are critical developmental windows for both the
microbiome and the brain
\cite{laueDevelopingMicrobiomeBirth2022}.
Fetal development is believed to occur in a sterile environment, but newborns
are rapidly seeded at birth through contact with the birth canal (if
birthed vaginally), caregivers, food (breast milk or formula),
and other environmental sources
\cite{backhedDynamicsStabilizationHuman2015,bokulichAntibioticsBirthMode2016}.
The early microbiome is characterized by low microbial
diversity, rapid succession and evolution, and is dominated by
Actinobacteria, particularly the genus \emph{Bifidobacterium},
Bacteroidetes, especially \emph{Bacteroides}, and Proteobacteria
\cite{koenigSuccessionMicrobialConsortia2011}.
Many of these bacteria have specialized metabolic capabilities
for digesting human milk, such as \emph{Bifidobacterium longum}
subsp. \textit{infantis} and \emph{Bacteroides fragilis}
\cite{selaGenomeSequenceBifidobacterium2008}.
Upon the introduction of solid foods, the gut
microbiome undergoes another categorical transformation;
its diversity increases, and most taxa of the infant microbiome
are replaced by taxa more reminiscent of adult microbiomes
\cite{backhedDynamicsStabilizationHuman2015}.
Prior studies have typically focused on either infant microbiomes or
adult microbiomes, since performing statistical analyses across this
transition poses particular challenges. Nevertheless, since this
transition coincides with critical neural developmental windows and
associated neurodevelopmental processes including myelination, neurogenesis, and synaptic pruning,
investigation across this solid-food boundary is important
\cite{tauNormalDevelopmentBrain2010}.

@@

As several prior studies have demonstrated associations between specific gut taxa
and atypical neurocognition and neurological disorders
\cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021,
      magnussonRelationshipsDietrelatedChanges2015,mayneris-perxachsMicrobiotaAlterationsProline2022,
      needhamGutderivedMetaboliteAlters2022},
we sought to determine if specific taxa or gene functions were associated
with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism were associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the $\beta$ diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we separately considered ages that are generally pre-transition (prior to 6 months of age old)
and those that are generally post-transition (over 18 months of age).
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (Figure 1G).
We also found that overall variation in microbial species in children over 18 months old
was significantly associated with variation in cognitive function score
(Figure 1G, R\textsuperscript{2} = 0.0099, q \textless{} 0.01),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0115, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants (Figure 1G).
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0207, and  gene functions
annotated with UniRef90s clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0258, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging (MRI) profiles (R\textsuperscript{2} = 0.258, q
\textless{} 0.001).

@@

\begin{sidewaysfigure}[!htb]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure2.png}
    \caption{
        Taxa and gene functional groups are associated with cognitive function.
        (\textbf{A}) Volcano plot of multivariable linear model results showing the
        relationship between individual taxa and cognitive function score in
        children over 18 months of age (left) or across all ages (right),
        controlling for age, maternal education, and sequencing depth.
        Taxa that were significant after FDR correction (q
        \textless{} 0.2) are colored red. (\textbf{B}) For taxa that were significantly
        associated with cognitive function,
        heatmaps of correlation with cognitive function in different age groups.
        (\textbf{C}) Multivariate linear models showing the relationship
        between individual taxa and domain-specific skill assessments
        from the MSEL for children under 6m (left), 18m to 36m (right),
        or over across all ages (right)
        controlling for age, maternal education, and sequencing depth.
        Asterisks in (A) and (C) indicate significance (q \textless{} 0.2)
        after multiple hypothesis correction.
        (\textbf{D-E}) Enrichment plots for proprionate degradation I gene set
        used in feature set enrichment analysis (FSEA) in children under 6m (D)
        or over 18m (E). (\textbf{G}) summary of FSEA
        results in children under 6m of age (left)
        over 18 months (middle) or across all ages (right), colored based on the
        significance of the association. Markers indicate the individual
        correlation of genes within a gene set,
        and vertical bars colored bars represent the
        median correlation of that gene set.
        Vertical dotted lines within each panel represent the median T statistic
        for all genes in that age group.
    }
    \label{fig:2}
\end{sidewaysfigure}

Cognitive assessments (MSEL, WPPSI, WISC)
are composites of multiple subscales that assess different cognitive domains.
While the composite scores are broadly analagous accross ages and test modality,
the subscales assessed in each test are different.
To assess whether microbial taxa were associated with
the development of specific cognitive domains,
we fit multivariate linear models for each species present in at least 10\%
of samples for children under 6 months, 18 to 36 months, or all ages
with a concurrent MSEL score (Figure 2C).
Several taxa associated with composite scores (Figure 2B)
were also associated with one or more subscores, including
\textit{F. prausnitzii} and \textit{A. obesi}.
Interestingly, while no taxa were significantly associated
with composite scores in chilren under 6 months old,
two species from the genus \textit{Streptococcus}
(\textit{S. peroris} and \textit{S. mitis}), as well as
\textit{Bacteroides fragilis}, were positively associated
with expressive language in this age group.
\textit{S. mitis} was also associated with expressive language
when looking at all ages, while \textit{Eisenbergiella massiliensis},
\textit{Anaerotruncus colihominis}, and an unclassified \textit{Blautia}
species were negatively associated. % Add more here if they match brain?

@@

\begin{table}[!h]
    \begin{center}
    \begin{tabular}{|r|r|r|r|r|}
      \hline\hline
      \textbf{Subject Ages (months)} & \textbf{Microbial feature} & \textbf{Demo.} &  \textbf{Test set correlation (± C.I.)} & \textbf{Test set RMSE (± C.I.)} \\\hline
      \multirow{5}{*}{0 to 6} & -    & + & 13.37 ± 0.05 & -0.16 ± 0.01 \\ \cline{2-5}
            & \multirow{2}{*}{taxa}  & - & 12.99 ± 0.04 & -0.11 ± 0.01 \\ \cline{3-5}
            &                        & + & 13.01 ± 0.04 & -0.13 ± 0.01 \\ \cline{2-5}
            & \multirow{2}{*}{genes} & - & 12.95 ± 0.04 & -0.05 ± 0.01 \\ \cline{3-5}
            &                        & + & 12.95 ± 0.04 & -0.05 ± 0.01 \\ \cline{1-5}
      \multirow{5}{*}{18 to 120} & - & + & 17.06 ± 0.04 & 0.511 ± 0.003 \\ \cline{2-5}
            & \multirow{2}{*}{taxa}  & - & 18.71 ± 0.04 & 0.347 ± 0.003 \\ \cline{3-5}
            &                        & + & 18.25 ± 0.04 & 0.429 ± 0.003 \\ \cline{2-5}
            & \multirow{2}{*}{genes} & - & 18.96 ± 0.04 & 0.299 ± 0.003 \\ \cline{3-5}
            &                        & + & 18.74 ± 0.04 & 0.341 ± 0.003 \\ \hline\hline
    \end{tabular}
    \caption{\label{tab:rfbench}Benchmark metrics for the cognitive assessment score
    prediction models. Confidence intervals (C.I.) are calculated from the
    distribution of metrics from repeated CV at a confidence level of 95.
    Root mean square error (RMSE) was used to estimate how well the random forest model was able to predict the outcome.
    Demo. column indicates whether demographic factors such as maternal education, sex and age were included as predictors.}
    \end{center}
\end{table}

\begin{table}[!h]
    \begin{center}
    \begin{tabular}{|r|r|r|r|r}
        \hline
        \textbf{Microbial feature} & \textbf{Demo.} & \textbf{Test set correlation (± C.I.)} & \textbf{Test set Root-mean-square error (± C.I.)} \\\hline
        - & + & 20.25 ± 0.07 & 0.43 ± 0.0 \\ \cline{1-4}
        \multirow{2}{*}{taxa} & - & 22.32 ± 0.08 & 0.07 ± 0.0 \\ \cline{2-4}
            & + & 21.53 ± 0.08 & 0.27 ± 0.0 \\ \cline{1-4}
        \multirow{2}{*}{genes} & - & 22.21 ± 0.08 & 0.12 ± 0.0 \\ \cline{2-4}
            & + & 22.08 ± 0.08 & 0.15 ± 0.0 \\\hline\hline
    \end{tabular}
    \caption{\label{tab:rfbench2}Benchmark metrics for the cognitive assessment score
    future prediction models. Confidence intervals (C.I.) are calculated from the
    distribution of metrics from repeated CV at a confidence level of 95.
    Root mean square error (RMSE) was used to estimate how well the random forest model was able to predict the outcome.
    Demo. column indicates whether demographic factors such as maternal education, sex and age were included as predictors.}
    \end{center}
\end{table}

@@

Nearly all important taxa in RF models that predicted cognitive function
in children older than 18 months were also important in
models that predicted future cognitive function, if they were present (Figure 3B).
At the same time, many additional species were important for future prediction models
that were not important in the models measuring concurrent cognitive function.
In particular, feature importance for future prediction models
were dominated by two species of \textit{Klebsiella},
\textit{K. pneumoniae} and \textit{K. varicola} (Figure 3D, E).
Many of the other species that ranked highest in feature importance
were also oportunistic pathogens, or previously reported
to be linked with disease (eg. \textit{R. gnavus}, \textit{E. coli}).

\subsubsection*{Gut microbial taxa are associated with specific cognitive functions} %TODO: add numbers!

Given that the development of different domains of cognitive function
(e.g., gross motor development vs visual perception) may develop at different times
and be affected by different molecular signals,
we sought to determine whether the associations of microbial taxa
with cognitive function are general across all cognitive subdomains.
To accomplish this, we assessed individual subscales of the  
Mullen Scales of Early Learning (MSEL, birth to 36 months - \textit{see Methods}). 
Taxonomic profiles coupled with demographics exhibited predictive potential towards three
of five probed subscales (Figure 4A), 
namely Expressive Language ( test-set RMSE 12.05, test-set correlation 0.368), 
Gross Motor (test-set RMSE 11.07, test-set correlation 0.137) and 
Visual Reception (test-set RMSE 15.2, test-set correlation 0.333).
Between the subscales, not only the predictability (measured by the test-set figures of merit) is different,
but also the relative importances of several taxa (Figure 4B).
Although certain species such as \textit{B. longum}, \textit{F. prausznitzii} 
have similar relative importance across all subscales, others are substantially different.
Notably, \textit{Bifidobacterium pseudocatenulatum}, \textit{B. wexlerae} and \textit{E. eligens} are
substantially more important for the prediction of EL, while
\textit{R. faecis}, \textit{Streptococcus salivarius} and \textit{F. saccharivorans} are
distinctly predictive of GM, and
\textit{Clostridium innocuum}, \textit{B. vulgatus} 
stand out on models targeting visual reception.

@@

Given that individual microbes could be associated with particular domains
of cognitive function and not others, if there are causal effects of microbial metabolism on cognitive
function, they might be reflected in changes in neuroanatomy. We again
used RF modeling to associate gut taxonomic
profiles with individual brain regions identified in MRI scans
(children over 18 months old, N = 124).
Some brain regions were more readily
predicted by RF models trained on microbial taxa (Table 4, Table S2), in particular those that were highly correlated with age.
These included the L/R lingual gyrus (mean RF correlations, Left =
0.444, Right = 0.441; importance for age, Left = 6.8\%, Right =
7.7\%) and the L/R pericalcarine cortex (mean RF correlations, Left =
0.267, Right = 0.315; importance for age, Left = 3.6\%, Right = 6.2\%).
In many cases, however, age was not an important variable in
high-performing models, such as that for the left accumbens area (mean
RF correlations = 0.275; relative age importances = 1.1\%).

>>>>

\begin{abstract}
The gastrointestinal tract, its resident microorganisms, and the central
nervous system are connected by biochemical signaling, also known as the
"microbiome-gut-brain-axis." Both the human brain and the gut microbiome
have critical developmental windows in the first years of life,
raising the possibility that their development is co-occurring and
likely co-dependent. Emerging evidence implicates gut microorganisms and
microbiota composition in cognitive outcomes and neurodevelopmental
disorders (e.g., autism and anxiety), but the influence of gut microbial
metabolism on typical neurodevelopment has not been explored in detail.
We investigated the relationship of the microbiome with the neuroanatomy
and cognitive function of 381 healthy children, demonstrating that
differences in gut microbial taxa and gene functions are associated with
overall cognitive function and with differences in the size of
multiple brain regions.
Using a combination of multivariate linear and machine learning (ML) models,
we showed that many species,
including \emph{Alistipes obesi} and \emph{Blautia wexlerae}, were % check RF models for direction of B. wex
associated with higher cognitive function, while some
species such as \emph{Ruminococcus gnavus} were more commonly found in % check RF models for direction of R. gnav
children with low cognitive scores after controlling for sociodemographic factors.
Microbial genes for enzymes involved in the metabolism of
neuroactive compounds, particularly short-chain fatty acids such as
acetate and propionate, were also associated with cognitive function.
In addition, ML models were able to use microbial taxa to predict the
volume of brain regions, and many taxa that were identified
as important in predicting cognitive function also dominated the
feature importance metric for individual brain regions.
For example, \emph{B. wexlerae} % is this still true?
was the most important species in models predicting the size of the parahippocampal region
in both the left and right hemispheres, while several species from
the phylum \textit{Bacteroidetes}, including GABA-producing % is this still true?
\emph{B. ovatus}, were important for predicting
the size of the left accumbens area, but not the right. These
findings provide potential biomarkers of neurocognition and brain development
and may lead to the future development of targets for early detection and early
intervention.
\end{abstract}

@@

The first years of life are critical developmental windows for both the
microbiome and the brain
\cite{laueDevelopingMicrobiomeBirth2022}.
Fetal development is believed to occur in a sterile environment, but newborns
are rapidly seeded at birth through contact with the birth canal (if
birthed vaginally), caregivers, food (breast milk or formula),
and other environmental sources
\cite{backhedDynamicsStabilizationHuman2015,bokulichAntibioticsBirthMode2016}.
The early microbiome is characterized by low microbial
diversity, rapid succession and evolution, and domination by
Actinobacteria, particularly the genus \emph{Bifidobacterium},
Bacteroidetes, especially \emph{Bacteroides}, and Proteobacteria
\cite{koenigSuccessionMicrobialConsortia2011}.
Many of these bacteria have specialized metabolic capabilities
for digesting human milk, such as \emph{Bifidobacterium longum}
subsp. \textit{infantis} and \emph{Bacteroides fragilis}
\cite{selaGenomeSequenceBifidobacterium2008}.
Upon the introduction of solid foods, the gut
microbiome undergoes another categorical transformation;
its diversity increases, and most taxa of the infant microbiome
are replaced by taxa more reminiscent of adult microbiomes
\cite{backhedDynamicsStabilizationHuman2015}.
Prior studies have typically focused on either infant microbiomes or
adult microbiomes, since performing statistical analyses across this
transition poses particular challenges. Nevertheless, since this
transition coincides with critical neural developmental windows and
associated neurodevelopmental processes including myelination, neurogenesis, and synaptic pruning,
investigation across this solid-food boundary is important
\cite{tauNormalDevelopmentBrain2010}.

@@

As several prior studies have demonstrated associations between specific gut taxa
and atypical neurocognition and neurological disorders
\cite{liuAlteredGutMicrobiota2019,wanUnderdevelopmentGutMicrobiota2021,
      magnussonRelationshipsDietrelatedChanges2015,mayneris-perxachsMicrobiotaAlterationsProline2022,
      needhamGutderivedMetaboliteAlters2022},
we sought to determine if specific taxa or gene functions were associated
with normal cognitive development in children.
To test if variations in gut microbial taxa,
their genes, or their metabolism were associated with neurocognitive
development, we used permutation analysis of variance (PERMANOVA) 
that included the $\beta$ diversity of microbial taxa and gene functions,
neuroimaging-derived volume measures of cortical and subcortical structures,
and general cognitive ability.
Given the large ecological shift in the microbiome that occurs upon the introduction of solid food,
and the relatively wide range of ages when over which infant transition from milk to solid foods,
we separately considered ages that are generally pre-transition (prior to 6 months of age old)
and those that are generally post-transition (over 18 months of age).
However, even with this categorization, we find increasing variation
in gut microbiomes in children over 18 months old (Figure 1G).
We also found that overall variation in microbial species in children over 18 months old
was significantly associated with variation in cognitive function score
(Figure 1G, R\textsuperscript{2} = 0.0099, q \textless{} 0.01),
as was variation in microbial gene functions (R\textsuperscript{2} = 0.0119, q \textless{} 0.001).
Variation in microbial taxa and genes was not significantly associated with cognitive
function in children under 6 months, though this may be due to the low
taxonomic diversity and broad lack of overlap between taxa in infants (Figure 1G).
As expected, age was significantly associated with microbial beta
diversity (taxa R\textsuperscript{2} = 0.0207, and  gene functions
annotated with UniRef90s clusters of 90\% similarity
\cite{suzekUniRefComprehensiveNonredundant2007} -
R\textsuperscript{2} = 0.0263, q \textless{} 0.001), and very strongly
associated with variation in neuroimaging (MRI) profiles (R\textsuperscript{2} = 0.258, q
\textless{} 0.001).

@@

\begin{sidewaysfigure}[!htb]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure2.png}
    \caption{
        Taxa and gene functional groups are associated with cognitive function.
        (\textbf{A}) Volcano plot of multivariable linear model results showing the
        relationship between individual taxa and cognitive function score in
        children over 18 months of age (left) or across all ages (right),
        controlling for age, maternal education, and sequencing depth.
        Taxa that were significant after FDR correction (q
        \textless{} 0.2) are colored red. (\textbf{B}) For taxa that were significantly
        associated with cognitive function,
        heatmaps of correlation with cognitive function in different age groups.
        (\textbf{C}) Multivariate linear models showing the relationship
        between individual taxa and domain-specific skill assessments
        from the MSEL for children under 6m (left), 18m to 48m (right),
        or over across all ages (right)
        controlling for age, maternal education, and sequencing depth.
        Asterisks in (A) and (C) indicate significance (q \textless{} 0.2)
        after multiple hypothesis correction.
        (\textbf{D-E}) Enrichment plots for proprionate degradation I gene set
        used in feature set enrichment analysis (FSEA) in children under 6m (D)
        or over 18m (E). (\textbf{G}) summary of FSEA
        results in children under 6m of age (left)
        over 18 months (middle) or across all ages (right), colored based on the
        significance of the association. Markers indicate the individual
        correlation of genes within a gene set,
        and vertical bars colored bars represent the
        median correlation of that gene set.
        Vertical dotted lines within each panel represent the median T statistic
        for all genes in that age group.
    }
    \label{fig:2}
\end{sidewaysfigure}

Cognitive assessments (MSEL, WPPSI, WISC)
are composites of multiple subscales that assess different cognitive domains.
While the composite scores are broadly analagous accross ages and test modality,
the subscales assessed in each test are different.
To assess whether microbial taxa were associated with
the development of specific cognitive domains,
we fit multivariate linear models for each species present in at least 10\%
of samples for children under 6 months, 18 to 48 months, or all ages
with a concurrent MSEL score (Figure 2C).
Several taxa associated with composite scores (Figure 2B)
were also associated with one or more subscores, including
\textit{F. prausnitzii} and \textit{A. obesi}.
Interestingly, while no taxa were significantly associated
with composite scores in chilren under 6 months old,
two species from the genus \textit{Streptococcus}
(\textit{S. peroris} and \textit{S. mitis}), as well as
\textit{Bacteroides fragilis}, were positively associated
with expressive language in this age group.
\textit{S. mitis} was also associated with expressive language
when looking at all ages, while \textit{Eisenbergiella massiliensis},
\textit{Anaerotruncus colihominis}, and an unclassified \textit{Blautia}
species were negatively associated. % Add more here if they match brain?

@@

\begin{table}[!h]
    \begin{center}
    \begin{tabular}{|r|r|r|r|r|}
      \hline\hline
      \textbf{Subject Ages (months)} & \textbf{Microbial feature} & \textbf{Demo.} &  \textbf{Test set RMSE (± C.I.)} & \textbf{Test set correlation (± C.I.)} \\\hline
      \multirow{5}{*}{0 to 6} & -    & + & 13.37 ± 0.05 & -0.16 ± 0.01 \\ \cline{2-5}
            & \multirow{2}{*}{taxa}  & - & 12.99 ± 0.04 & -0.11 ± 0.01 \\ \cline{3-5}
            &                        & + & 13.01 ± 0.04 & -0.13 ± 0.01 \\ \cline{2-5}
            & \multirow{2}{*}{genes} & - & 12.95 ± 0.04 & -0.05 ± 0.01 \\ \cline{3-5}
            &                        & + & 12.95 ± 0.04 & -0.05 ± 0.01 \\ \cline{1-5}
      \multirow{5}{*}{18 to 120} & - & + & 17.06 ± 0.04 & 0.511 ± 0.003 \\ \cline{2-5}
            & \multirow{2}{*}{taxa}  & - & 18.71 ± 0.04 & 0.347 ± 0.003 \\ \cline{3-5}
            &                        & + & 18.25 ± 0.04 & 0.429 ± 0.003 \\ \cline{2-5}
            & \multirow{2}{*}{genes} & - & 18.96 ± 0.04 & 0.299 ± 0.003 \\ \cline{3-5}
            &                        & + & 18.74 ± 0.04 & 0.341 ± 0.003 \\ \hline\hline
    \end{tabular}
    \caption{\label{tab:rfbench}Benchmark metrics for the cognitive assessment score
    prediction models. Confidence intervals (C.I.) are calculated from the
    distribution of metrics from repeated CV at a confidence level of 95.
    Root mean square error (RMSE) was used to estimate how well the random forest model was able to predict the outcome.
    Demo. column indicates whether demographic factors such as maternal education, sex and age were included as predictors.}
    \end{center}
\end{table}

@@

\begin{table}[!h]
    \begin{center}
    \begin{tabular}{|r|r|r|r|r}
        \hline
        \textbf{Microbial feature} & \textbf{Demo.} &  \textbf{Test set Root-mean-square-error (± C.I.)} & \textbf{Test set correlation (± C.I.)} \\\hline
        - & + & 20.25 ± 0.07 & 0.43 ± 0.0 \\ \cline{1-4}
        \multirow{2}{*}{taxa} & - & 22.32 ± 0.08 & 0.07 ± 0.0 \\ \cline{2-4}
            & + & 21.53 ± 0.08 & 0.27 ± 0.0 \\ \cline{1-4}
        \multirow{2}{*}{genes} & - & 22.21 ± 0.08 & 0.12 ± 0.0 \\ \cline{2-4}
            & + & 22.08 ± 0.08 & 0.15 ± 0.0 \\\hline\hline
    \end{tabular}
    \caption{\label{tab:rfbench2}Benchmark metrics for the cognitive assessment score
    future prediction models. Confidence intervals (C.I.) are calculated from the
    distribution of metrics from repeated CV at a confidence level of 95.
    Root mean square error (RMSE) was used to estimate how well the random forest model was able to predict the outcome.
    Demo. column indicates whether demographic factors such as maternal education, sex and age were included as predictors.}
    \end{center}
\end{table}

Nearly all important taxa in RF models that predicted cognitive function
in children older than 18 months were also important in
models that predicted future cognitive function, if they were present (Figure 3B, Table S5).
At the same time, many additional species were important for future prediction models
that were not important in the models measuring concurrent cognitive function.
In particular, feature importance for future prediction models
were dominated by two species of \textit{Klebsiella},
\textit{K. pneumoniae} and \textit{K. varicola} (Figure 3D, E).
Many of the other species that ranked highest in feature importance
were also oportunistic pathogens, or previously reported
to be linked with disease (eg. \textit{R. gnavus}, \textit{E. coli}).

\subsubsection*{Gut microbial taxa are associated with specific cognitive functions}

Given that the development of different domains of cognitive function
(e.g., gross motor development vs visual perception) may develop at different times
and be affected by different molecular signals,
we sought to determine whether the associations of microbial taxa
with cognitive function are general across all cognitive subdomains.
To accomplish this, we assessed individual subscales of the  
Mullen Scales of Early Learning (MSEL, birth to 36 months - \textit{see Methods}). 
Taxonomic profiles coupled with demographics exhibited predictive potential towards three
of five probed subscales (Figure 4A, Table 4), 
namely Expressive Language ( test-set RMSE 12.05, test-set correlation 0.368), 
Gross Motor (test-set RMSE 11.07, test-set correlation 0.137) and 
Visual Reception (test-set RMSE 15.2, test-set correlation 0.333).
Between the subscales, not only the predictability (measured by the test-set figures of merit) is different,
but also the relative importances of several taxa (Figure 4B).
Although certain species such as \textit{B. longum}, \textit{F. prausznitzii} 
have similar relative importance across all subscales, others are substantially different (Tables S2-S4).
Notably, \textit{Bifidobacterium pseudocatenulatum}, \textit{B. wexlerae} and \textit{E. eligens} are
substantially more important for the prediction of EL, while
\textit{R. faecis}, \textit{Streptococcus salivarius} and \textit{F. saccharivorans} are
distinctly predictive of GM, and
\textit{Clostridium innocuum}, \textit{B. vulgatus} 
stand out on models targeting visual reception.

\begin{table}[!h]
    \begin{center}
        \begin{tabular}{|r|r|r|r|r|}
        \hline
        \textbf{Target Mullen scale} & \textbf{Microbial feature} & \textbf{Demo.} & \textbf{Test set RMSE (± C.I.)} & \textbf{Test set correlation (± C.I.)} \\\hline
        \multirow{5}{*}{Expressive Language} & - & + & 11.17 ± 0.04 & 0.48 ± 0.0 \\ \cline{2-5}
            & \multirow{2}{*}{taxa}  & - & 12.37 ± 0.04 & 0.25 ± 0.0 \\ \cline{3-5}
            &                        & + & 12.05 ± 0.04 & 0.37 ± 0.0 \\ \cline{2-5}
            & \multirow{2}{*}{genes} & - & 12.59 ± 0.04 & 0.15 ± 0.0 \\ \cline{3-5}
            &                        & + & 12.46 ± 0.04 & 0.21 ± 0.0 \\ \cline{1-5}
        \multirow{5}{*}{Gross Motor} & - & + & 10.53 ± 0.04 & 0.346 ± 0.006 \\ \cline{2-5}
            & \multirow{2}{*}{taxa}  & - & 11.08 ± 0.04 & 0.134 ± 0.006 \\ \cline{3-5}
            &                        & + & 11.07 ± 0.04 & 0.137 ± 0.006 \\ \cline{2-5}
            & \multirow{2}{*}{genes} & - & 10.98 ± 0.04 & 0.197 ± 0.005 \\ \cline{3-5}
            &                        & + & 10.98 ± 0.04 & 0.197 ± 0.005 \\ \hline\hline
        \multirow{5}{*}{Visual reception} & - & + & 14.13 ± 0.04 & 0.462 ± 0.003 \\ \cline{2-5}
            & \multirow{2}{*}{taxa}  & - & 15.49 ± 0.04 & 0.245 ± 0.004 \\ \cline{3-5}
            &                        & + & 15.2 ± 0.04 & 0.333 ± 0.004 \\ \cline{2-5}
            & \multirow{2}{*}{genes} & - & 15.97 ± 0.04 & 0.067 ± 0.004 \\ \cline{3-5}
            &                        & + & 15.84 ± 0.04 & 0.118 ± 0.005 \\ \hline\hline
        \end{tabular}
        \caption{\label{tab:subscalebenchmark}Benchmark metrics for the Mullen subscales prediction models.
        Confidence intervals (C.I.) are calculated from the
        distribution of metrics from repeated CV at a confidence level of 95.
        Root mean square error (RMSE) was used to estimate how well the random forest model was able to predict the outcome.
        Demo. column indicates whether demographic factors such as maternal education, sex and age were included as predictors.
        Age range, in all cases, was 18-120 months.}
    \end{center}
\end{table}

@@

Given that individual microbes could be associated with particular domains
of cognitive function and not others, if there are causal effects of microbial metabolism on cognitive
function, they might be reflected in changes in neuroanatomy. We again
used RF modeling to associate gut taxonomic
profiles with individual brain regions identified in MRI scans
(children over 18 months old, N = 124).
Some brain regions were more readily
predicted by RF models trained on microbial taxa (Table 5, Table S2), in particular those that were highly correlated with age.
These included the L/R lingual gyrus (mean RF correlations, Left =
0.444, Right = 0.441; importance for age, Left = 6.8\%, Right =
7.7\%) and the L/R pericalcarine cortex (mean RF correlations, Left =
0.267, Right = 0.315; importance for age, Left = 3.6\%, Right = 6.2\%).
In many cases, however, age was not an important variable in
high-performing models, such as that for the left accumbens area (mean
RF correlations = 0.275; relative age importances = 1.1\%).
